\chapter{Image processing menus}
\mylabel{sec:menus}

\noindent
This chapter is runs quickly through the \ct{Toolkits} menu.  See
\cref{sec:program} if you want to understand how the menus are written
(or want to add more of your own). Use the Toolkit Browser to find stuff.

Some things are common to almost all menu items:

\begin{description}

\item[Tooltips]
	If you rest your mouse pointer over an item,
	you'll see a quick description of what the item does.

\item[Grouping]
	You can select several objects, click \ctr{Edit}\ct{Group}, and then
	when you click the item, it will operate on all the objects in the 
	group.

\item[Any type]
	Almost all items will work on any object. You can add an image and a
	number, for example, find the colour difference between a number and
	an image, or transform a matrix from LAB to XYZ.

\end{description}

\section{Colour}
\mylabel{sec:menu-colour}

This menu groups operations on colorimetric images and patches of
colour. A colour patch is three float numbers plus a tag saying how those
number should be interpreted as colour (for example, as a colour in CIE LAB
colourspace). You can drag and drop between colour patches, and into and from
the inkwell in an image paint window. Double-left-click on a colour patch to
open a colour select dialog.

\nip{} has 9 main types of colorimetric image, see \tref{tb:colour}.
All these types are D65 (that is, daylight) absolute colorimetric.
When it displays an image, \nip{} uses the \ct{Type} field in the image
header as a hint on how to transform the numbers in the image into RGB for
the display. The current \ct{Type} is displayed at the end of the caption
line below an image thumbnail. 

\begin{tab2}
\begin{center}
\begin{tabular}{||l|l|l||}
\hline
Name & Format & Notes \\
\hline
\ct{Mono} & One band 8 bit & 
	Not really calibrated at all \\
\ct{sRGB} & Three band 8 bit & 
	Screen device space for the sRGB standard \\
\ct{Lab} & Three band float & 
	The 1976 version of the CIE perceptual colourspace \\
\ct{LabQ} & Four band 8 bit & 
	Like \ct{Lab}, but represented as 10:11:11 bits \\
\ct{LabS} & Three band 16 bit & 
	Like \ct{Lab}, but represented as 15:16:16 bits \\
\ct{LCh} & Three band float & 
	\ct{Lab}, but with polar coordinates \\
\ct{XYZ} & Three band float & 
	The base CIE colourspace \\
\ct{Yxy} & Three band float & 
	Sometimes useful for colour meters \\
\ct{UCS} & Three band float & 
	Highly uniform space from the CMC(l:c) standard \\
\hline
\end{tabular}
\end{center}
\caption{\nip{} colourspaces}
\mylabel{tb:colour}
\end{tab2}

\begin{description}

\item[\ct{New}]
	Make a patch of colour, or pick a colour from a slice through CIELAB
	colourspace.

\item[\ct{Convert To Colour}]
	Convert anything into a Colour object.

\item[\ct{Colourspace}]
	Change the colourspace. The stored numbers change, but the
	visual appearance should stay the same.

\item[\ct{Tag As}]
	Change the colourspace tag (the \ct{Type} field in the image
	header). The stored numbers stay the same, but the visual appearance
	should change.

\item[\ct{Colour Temperature}]
	Change the colour temperature. \ct{Move Whitepoint} just adjusts the
	ratios of X and Z using the CIE standard illuminants.
	
	\ct{D65 to D50} and \ct{D50 to D65} transform using either a 3x3
	matrix which is numerically minimal in XYZ space with respect to the
	colours on a Macbeth Color Checker, or via Bradford cone space. The
	Bradford transform omits the power term.

	The final two items go from XYZ to LAB and back, but with D50
	normalisation rather than the default D65.

\item[\ct{ICC}]
	Transform images (not patches of colour) device space to profile 
	connection space (LAB float) and back. 
	
	You need to be careful about colour temperature issues:
	all printers work with D50, and \nip{} is all D65. Use the D65 to D50
	interchange items in the \ct{Colour Temperature} menu to swap back and
	forth.

	All printers also work with relative colorimetry, and \nip{} is
	generally absolute. Use \ct{Absolute to Relative} to scale an absolute
	colorimetric image by a media white point.

\item[\ct{Difference}]
	Calculate various colour difference metrics. You can mix patches of
	colour and colour images.

\item[\ct{Adjust}]
	Change colour in a colorimetric way. \ct{Recombination}
	multiplies each pixel in an image through a matrix. \ct{Cast}
	displaces the neutral axis in LAB space. \ct{HSB} lets you adjust an
	image in LCh colourspace.

\item[\ct{Similar Colour}]
	find pixels in an image with a similar colour to a patch of
	colour. 

\item[\ct{Measure Colour Chart}]
	This takes a trimmed image of a colour chart (a rectangular grid of
	coloured squares), measures the average pixel value in the centre 50\%
	of each square, and returns a matrix of the measured values. 

	Use \ct{Make Synthetic Colour Chart} to make a colour chart image from
	a matrix of measurements. 

\item[\ct{Plot ab Scatter}]
	draws a 2 dimensional histogram of the distribution of pixel colours
	in LAB colourspace.

\end{description}

\section{Filter}
\mylabel{sec:menu-filter}

This menu groups operations which filter images, or which are filters in the
photoshop sense.

\begin{description}

\item[\ct{Convolution}]
	This menu has several standard convolution operations (blur, sharpen,
	edge detect, etc.), plus the option to convolve with a custom kernel.

	Two menu items are slightly more complicated. \ct{Unsharp Mask}
	transforms to CIE LAB colour space, then sharpens just the L band
	with a cored unsharp filter. The \ctr{Tasks}\ct{Print} menu has a
	version of this filter tuned for typical inkjet printers.

	\ct{Custom Blur} builds and applies a square or gaussian convolution
	kernel for you based on a radius setting.

\item[\ct{Rank}]
	A preset median filter, and a custom rank filter that lets you specify
	window size and rank.

	The \ct{Image Rank} item does pixel-wise ranking of a set of images.

\item[\ct{Morphology}]
	These menu items implement basic morphological operations. Images are
	zero for background and non-zero (usually 255) for object. Matricies
	are shown as 0, 1 and * for background, object and don't-care.

	The \ct{Threshold} item does a simple level threshold. Use the
	\ctr{Math}\ct{Relational} menu to construct more complex image
	binarisations. Use \ctr{Math}\ct{Boolean} to combine morphologies.

	The first half of the menu lists simple erode and dilate operations, 4-
	and 8-way connected. The second half contains several useful compound
	filters.

	See also \ctr{Histogram}\ct{Find Profile} for
	something that can search an image for object edges. And
	\ct{Math}\ctr{Statistics}\ct{Edges} can count the number of edges
	across and down an image.

\item[\ct{Fourier}]
	A selection of ideal, Gaussian and Butterworth Fourier space filters.

	You can make other mask shapes yourself using the \ctr{Image}\ct{Make
	Patterns} menus, then apply them using \ctr{Math}\ct{Fourier}. You can
	also use the image paintbox to directly paint out peaks in a
	fourier-space image before transforming back to real space.
	
\item[\ct{Enhance}]
	A selection of simple image enhancement filters. \ct{Statistical
	Difference} passes a window over an image and tries to match the
	region statistics at each point to a target mean and deviation. 

\item[\ct{Spatial Correlation}]
	Place a small image at every possible position in a big image and
	calculate the correlation at each position. \ct{Simple Difference} is
	the much faster unnormalised version.

\item[\ct{Tilt Brightness}]
	A selection of tools for adjusting the brightness of an image across
	it's surface. Useful for correcting lighting problems.

\item[\ct{Blend}]
	Blend two objects together using either a third object to control the
	blend at each point, or a slider to set all points together. You can
	blend almost anything with anything.

	One useful version is to use a text image (see \ctr{Image}\ctr{Make
	Patterns}\ct{Text}) to blend between two colours (see
	\ctr{Colour}\ct{New}).

	\ct{Along Line} does a left/right or top/bottom fade between two
	images.

\item[\ct{Overlay}]
	Make a colour overlay of two monochrome images. Useful with
	\ctr{Image}\ctr{Transform} for testing image superposition.

\item[\ct{Colourize}]
	Use a colour image to tint a monochrome image. Useful in conjunction
	with \ctr{Image}\ctr{Transform}.

\item[\ct{Browse}]
	Look at either the bits or the bands of an image.

\item[\ct{Photographic Negative} and friends]
	A small selection of faintly photoshop-style filters. 

\end{description}

\section{Format}
\mylabel{sec:menu-format}

This groups a few items which had no obvious home and which change the format
of objects.

\begin{description}

\item[\ct{Duplicate}]
	Copy an object, stripping off any derived classes. For images, this
	really takes a copy of the underlying object (using \ct{im\_copy()}).

\item[\ct{List to Group}]
	Changes lists (see \ctr{Math}\ct{List}) into Groups (see
	\ctr{Edit}\ct{Group}) and back. A list os an ordered collection of
	objects. A group is a list that \nip{} will automatically iterate
	over.

\item[\ct{Break Up Object}]
	This tries to take an object apart. So a multi-band image becomes a
	list of 1-band images. A matrix becomes a list of vectors, and so on.

\item[\ct{CSV Import}]
	Import an image from a CSV file, with a few controls.

\item[\ct{Interpret Analyze 7 Header}]
	Read the meta fields for volume layout and calibration from the
	Analyze header and reformat the image appropriately.

\end{description}

\section{Histogram}
\mylabel{sec:menu-histogram}

This menu groups operations for finding and transforming image histograms.
\nip{} represents histograms and lookup tables as images with \ct{Type}
set to \ct{Histogram}.  Histograms may have pixels in any format and any
number of bands. You can only find histograms of unsigned 8- and 16-bit
images.

\begin{description}

\item[\ct{New Histogram}]
	This makes a new ramp histogram. A set of sliders let you adjust the
	shape. Use \ct{Map Histogram} to apply your ramp to an image.

\item[\ct{Build LUT from Scatter}]
	Make a histogram from a set of $(x, y)$ values. 

\item[\ct{Tag Image as Histogram}]
	Mark an image as actually being a histogram after all.

\item[\ct{Find Histogram}]
	A one dimensional histogram treats each band as an independent
	variable. An $n$-dimensional histogram treats each pixel as a vector of
	$n$ elements, where $n$ is the number of bands in the image.

\item[\ct{Map Histogram}]
	Looks up each pixel in the input in the histogram and send the found
	value to the output.

\item[\ct{Equalise Histogram}]
	Find the global or locally histogram equalised image.

\item[\ct{Cumulative Histogram}]
	Use this and friends to calculate a cumulative histogram, normalise a
	histogram and match two histograms.

\item[\ct{Find Profile}]
	Searches from the edges of an image for the first non-zero pixel and
	returns a profile histogram. 

\item[\ct{Find Projections}]
	Sum columns and rows.

\item[\ct{Plot Slice}]
	Mark a guide on an image (drag from the
	image rulers, or click \ctr{File}\ctr{New}\ct{Guide}) and click 
	\ct{Plot Slice} to make a histogram which is a horizontal or vertical 
	slice through an image. Use \ct{Extract Arrow} to extract the area
	around an arrow or guide. Use \ct{Plot Object} to make a plot of any
	object.

\end{description}

\section{Image}
\mylabel{sec:menu-image}

This menu groups operations which apply only to images.

\begin{description}

\item[\ct{New}]
	Makes a new image. \ct{Region on Image} makes a new region, arrow,
	guide or mark on an image. It's usually easier to open a viewer on
	an image and Ctrl-drag.

\item[\ct{Convert to Image}]
	Try to make an image out of anything.

\item[\ct{Header}]
	Try to change or examine the image header in various ways.

\item[\ct{Set Format}]
	Switch between the various precisions.

\item[\ct{Levels}]
	A variety of things for adjusting the values of pixels globally.

\item[\ct{Levels}]
	Various tools that change the levels in an image. \ct{Tone Curve} is
	the only complex one: it lets you adjust the image levels with a set
	of sliders.

\item[\ct{Transform}]
	Various tools that change the geometry of an image. 

	To use \ctr{Rotate}\ct{Straighten}, mark an arrow on an image
	(Ctrl-drag up and left in an image view window) along a
	near-horizontal or near-vertical edge. When you click on
	\ctr{Rotate}\ct{Straighten}, \nip{} will rotate the image by the
	smallest amount that makes that edge exactly horizontal or vertical.

	\ct{Linear Match} takes two images and rotates and scales
	the second so that the images can be superimposed. Drag the
	tie-=points to mark common features. Use \ctr{Filter}\ct{Overlay}
	or \ctr{Filter}\ct{Colourize} to actually superimpose them.

	\ct{Rubber Sheet} is useful for fixing things like lens distortion.
	You give \ct{Find} two images, a reference and a distorted version
	of that reference, and it automatically finds a transform which
	will map the distorted image back on to the reference image. Use
	\ct{Apply} to apply the discovered transform to another image.

\item[\ct{Band}]
	Extract/insert/delete image bands. Use \ct{To Dimension} to change
	image bands into a horizontal or vertical dimension. Use \ct{To
	Bands} to compress the horizontal or vertical dimension into bands
	(small images only!).

\item[\ct{Crop}]
	Crops an image. It's often easier to drag out a region. This menu item
	is only really useful for cropping large groups of images.

\item[\ct{Insert}]
	This takes two images and pastes the smaller into the centre of the
	larger. The two images have to have the same number of bands.
	If you open an image viewer on the large image, you'll see an area
	which you can drag around to set the exact insert point.

\item[\ct{Select}]
	Draw elipses and polygons on an image. Useful for selecting defined
	areas.

\item[\ct{Join}]
	Use to join two images together bandswise, left/right or up/down. 
	\ct{Array}
	joins a list of lists of images together into a single large image.

\item[\ct{Tile}]
	Repeat an image horizontally and vertically to make a larger image, or
	chop an image into a set of tiles.

\item[\ct{Patterns}]
	These items all make useful images for you, from checkerboards to
	gaussian masks. \ct{XY Image} is the most useful: you can use it to
	build other patterns.

\item[\ct{Test Images}]
	These items make a variety of useful testcharts for evaluating
	spatial response and colour.

\end{description}

\section{Math}
\mylabel{sec:menu-math}

Basic maths operations on any combination of any objects. You can add a slider
to a matrix, for example, then divide by an image. Hopefully most of these 
are obvious.

\begin{description}

\item[\ctr{Arithmetic}\ct{Absolute Value Vector}]
	The absolute value item normally calculates mod of each band of an
	image separately. By contrast, \ct{Absolute Value Vector} treats
	each pixel as a vector and calculates the modulus of that.

\item[\ct{List}]
	These aren't really maths operations, but they're in here too.

\end{description}

\section{Matrix}
\mylabel{sec:menu-matrix}

This menu groups operations which operate on matricies. \nip{} has four ways
of displaying a matrix, but they all behave in the same way under the skin.
Almost all the items in the \ct{Math} menu will work on matricies. Most of the
matrix operations will also work on images.

\begin{description}

\item[\ct{New}]
	The first four items make matricies which display and edit in various 
	ways useful for different applications. The final two make matricies
	which are pre-filled with useful numbers.

\item[\ct{Convert to Matrix}]
	Try to make anything into a matrix.

\item[\ct{Extract}]
	This group of items extracts various submatricies. You can also do
	this graphically: just drag-select an area in matrix.

\item[\ct{Insert}, \ct{Delete}, \ldots]
	Also work on images, which can be handy.
	A 45 degree rotate will only work for square matricies with
	odd-length sides.

\item[\ct{Invert}]
	Simple matrix-only maths operations.

\item[\ct{Plot Scatter}]
	This takes a two-column matrix where the columns are the X and Y
	positions of points and draws a scatter graph.

\end{description}

\section{Tasks}
\mylabel{sec:menu-tasks}

This menu repeats many items from other menus, but tries to group them by
tasks they are useful for, rather than by function.

\subsection{Capture}
\mylabel{sec:menu-capture}

This menu groups operations which are useful in capturing images, or for the
initial processing you might want to do to an image captured from another
program.

\begin{description}

\item[\ct{Capture Video Frame}]
	This menu item will currently only work on Linux machines with a
	compatible video4linux capture card. See \pref{sec:vidpref}
	for notes on how it works.

\item[\ct{Smooth}]
	Use this to remove texture from images. It's handy in conjunction with
	\ct{Flatfield}.

\item[\ct{Flatfield}]
	Use this to correct homogeneity. Select an image of a piece of white
	(or mid-grey) card, then select the image to correct, then click
	\ct{Flatfield}. Use \ct{Smooth} to renmove texture from the white card
	if necessary.

	You can select a single white and a group of images to correct a large
	set in one step

\item[\ct{White Balance}]
	Use this to move the white point to make an area of the image you know
	to be white, white.
	Mark a region on an image, enclosing a patch you know to be white.
	Select the region and the image and click on \ct{White Balance}.

\item[\ct{Find Colour Calibration}]
	Use this to colour calibrate an image. Drag a region enclosing an
	image of a Macbeth Color Checker Chart and click \ct{Find Colour
	Calibration}. 

\item[\ct{Apply Colour Calibration}]
	Use this to apply the transform calculated by the previous item to
	another image. Select the calibration object, select the RGB image
	you want calibrated, and click \ct{Apply Colour Calibration}.

\end{description}

\subsection{Mosaic}
\mylabel{sec:menu-mosaic}

The items in this menu are discussed in appalling detail in
\cref{sec:ir}.

\begin{description}

\item[\ct{One Point}]
	Join two images left-right or top-bottom with a simple
	translation. Mark a point on each image to be joined (open image
	view window, Ctrl-left-click, drag to position), then click on the
	mosaic button. The
	operation performs elaborate tie-point adjustment, so your selection
	of a common feature does not have to be exact.

	The \ct{Manual} versions do not perform automatic tie-point correction
	and are useful when joing very difficult images.

\item[\ct{Two Point}]
	Do a join, but allow the right-hand (or bottom) image to rotate and
	scale if it will improve the match. You need to pick two points on
	each image.

\item[\ct{Balance}]
	Break a mosaic apart, examine average pixel value in the overlap
	regions, adjust brightness to match, and reassemble. This only works
	for images which have been produced just by mosaic joins! If you've
	done anything else to the image since loading it, the balance will
	fail with a mysterious message.

\item[\ct{Manual Balance}]
	Adjust the brightness in a set of masked areas to match. Useful for
	removing shadows.

\item[\ct{Rebuild}]
	Use this to mosaic up one set of files based on joins you made
	in another.  Breaks a mosaic part to component files, performs a
	string substitution on the file names, and reassembles. 

\item[\ct{Clone Area}]
	Select over- or under-exposed pixels in one image and replace them
	with the corresponding pixels from another image. Useful for removing
	lead numbers used to identify X-ray plates.

	The function
	operates on two 8-bit mono images.  Move and resize the region
	on the first image to define the area around the white number.
	Move the region on the second to overlapping area.  A section of the
	area on the second image is cloned and blended into the first image.
	The amount of the defined area to be cloned in defined by a slider
	within the output image.

\end{description}

\subsection{Picture Frame}
\mylabel{sec:menu-picture-frame}

Items useful for mocking up painting frames.

\subsection{Print}
\mylabel{sec:menu-print}

Items useful while preparing an image for printing.

\begin{description}

\item[\ct{Sharpen}]
	Sharpen an image for printing. This is a version of
	\ctr{Filter}\ctr{Convolution}\ct{Unsharp Mask} tuned for typical
	inkjet printers.

\item[\ct{Adjust Tone Curve}]
	Adjust the reproduction tone curve in LAB. Most useful for offset
	work, especially from transparencies.

\end{description}
